---
title: "Practical 2 - microbiota analyses"
subtitle: "2024 NUTRIM microbiome & metabolome workshop"
author: YOUR-NAME
date: last-modified
keep-md: false
embed-resources: true
format: html
theme: flatly
toc: true
---

## Instructions for practical 2

Follow the instructions online at:\
[david-barnett.github.io/2024-NUTRIM-microbiome/practical-2/web/practical2-instructions.html](https://david-barnett.github.io/2024-NUTRIM-microbiome/practical-2/web/practical2-instructions.html)

For practical 2, you will need to fill out this notebook by copying and modifying the instruction material.

Remember you can add code chunks with `Ctrl`+`Alt`+`I` (Windows) or `Cmd`+`Option`+`I` (Mac).

Remember to take your time to understand each step. Don't rush to the end, there is no prize for completion!

## Research questions

Do include your own notes. When you are working on your own analyses, you can share these documents with your supervisors and collaborators.

. . .

### Load R packages

```{r}

```

### Read phyloseq data

. . .

```{r}

```

. . .

## Alpha diversity analyses

. . .

### Richness

. . .

```{r}

```

. . .

### Diversity

. . .

```{r}

```

. . .

## Dissimilarity & Ordination

. . .

### PCoA plots

```{r}

```

. . .

## PERMANOVA

. . .

```{r}

```

. . .

## Interactive ordination!

Don't include interactive code in the notebook sections.

But do include some code for ordination plots that `ord_explore` helped you to create.

```{r}

```

. . .

## PCA

. . .

```{r}

```

. . .

### Circular bar chart

```{r}

```

## Differential abundance

### Bar charts

You could include some stacked bar charts here before the differential abundance testing.

```{r}

```

. . .

### Taxon regression models

. . .

```{r}

```

. . .

### Tree plots

. . .

```{r}

```

. . .

## Extension activities?

If you get to the extension activities, you can add that code to this document too.

Keep things organised with headings.

```{r}

```

. . .

## Session info

```{r}
sessioninfo::session_info()
```
