---
title: "Practical 1B - microbiota data"
subtitle: "NUTRIM microbiome & metabolome workshop"
author: David Barnett
date: last-modified
keep-md: false
theme: 
  light: flatly
  dark: darkly
format: html
embed-resources: true
code-block-border-left: true
code-block-bg: true
toc: true
toc-location: right
toc-depth: 4
toc-expand: 3
number-sections: true
number-depth: 3
fig-align: center
fig-dpi: 300
fig-width: 7.5
fig-height: 5
fig-responsive: true
code-tools: true
code-fold: false
code-link: true
lightbox: auto
cache: false
---

## Intro

Now we're going to look at the microbiota data.

::: {.callout-note collapse="true"}
### A note on sequence data processing

These data have already been processed into a table, counts of how often each sequence occurs in each sample.

It started as huge fastq text files, full of As, Cs, Ts and Gs, but we will not practice sequence data processing today, because it takes quite a long time to run.

If you generate your data with us at Medical Microbiology, we will run the latest processing approaches on our computational infrastructure and provide you with the processed data.

Nowadays we do amplicon sequencing with Illumina MiSeq, HiSeq, or similar technologies, and denoise the output with DADA2 to produce ASV count tables.

The example data we will use are a little older. The amplicons were sequenced with 454 pyrosequencing and clustered into OTUs, but the core principles of analysis remain the same.

And as you will see later today, the same approaches can also be applied to taxonomic abundance count tables obtained from shotgun metagenomic sequencing.
:::

First we're going to use the R skills we practised in part A to inspect this data.

::: {.callout-note collapse="true" appearance="simple"}
### SOLUTION: (read me?)

-   In this practical, solution blocks like this are available...

-   But try and write code independently before looking at the answer!

-   Attempting to recall knowledge or solve problems is proven to enhance learning.

-   So don't look unless you're really stuck! 💪
:::

After that, we'll take a look at some specialist R packages for microbiome analysis.

## Learning goals 💪

-   [blah](#blah)

## Load R packages 📦

```{r}
library(readxl)
library(here)
library(tidyverse)
```

## Read and inspect data 🔍

Read the metadata file from part 1A, using the RDS version.

```{r}
meta <- read_rds(file = here("data/papa2012/processed/all_metadata.rds"))
```

Read the count table: this is stored as a TSV (tab-separated variables) formatted text file.

```{r}
counts <- read_tsv(file = here("data/papa2012/papa2012_OTU_count_table.txt"))
```

```{r}
counts
```

**Your first challenge:** read the taxonomy table stored in `"data/papa2012/papa2012_taxonomy_table.txt"` (call the object `taxonomy`)

::: {.callout-note collapse="true" appearance="simple"}
#### SOLUTION:

```{r}
taxonomy <- read_tsv(file = here("data/papa2012/papa2012_taxonomy_table.txt"))
```

```{r}
taxonomy
```
:::

### Taxonomy table

Now practice inspecting the taxonomy table by completing the following tasks:

::: panel-tabset
##### Task 1

Check how many distinct genera there are. Tip: use `unique()` and `length()`

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

```{r}
taxonomy$Genus %>%
  unique() %>%
  length()
```
:::

##### Task 2

How many OTUs are there in each Phylum? Tip: use `count()` or `table()`

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

```{r}
taxonomy %>% count(Phylum)
```

```{r}
taxonomy$Phylum %>% table(useNA = "ifany")
```
:::

##### Task 3

What genera are in the phylum Actinobacteria? Tip: use `filter()`

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

```{r}
taxonomy %>%
  filter(Phylum == "Actinobacteria") %>%
  count(Genus, sort = TRUE)
```

Or for just their names:

```{r}
taxonomy %>%
  filter(Phylum == "Actinobacteria") %>%
  pull(Genus) %>%
  unique()
```
:::
:::

### OTU count table

First let's plot a histogram of OTU number 1.

```{r, fig.width=8, fig.height=2}
counts %>% ggplot(aes(OTU_00001)) +
  geom_histogram(binwidth = 5)
```

Looks like there are a lot of zeros!

```{r}
table(OTU00001_has_0_counts = counts$OTU_00001 == 0, useNA = "ifany")
```

Attempt the following tasks to explore further!

::: panel-tabset
##### Task 1

Use `filter()` to plot only the non-zero entries for OTU 1.

Try also using `+ scale_x_log10()` to transform the plot axis scale.

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

```{r, fig.width=8, fig.height=2}
counts %>%
  filter(OTU_00001 != 0) %>%
  ggplot(aes(OTU_00001)) +
  geom_histogram(bins = 30) +
  scale_x_log10()
```
:::

##### Task 2

Use `mutate()` to temporarily create a new log-transformed variable called log_OTU1 and plot its distribution.

*Note that because you cannot log-transform 0s you will need to add 1 to all values first, i.e.* `log10(OTU_00001 + 1)`

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

```{r, fig.width=8, fig.height=2}
counts %>%
  mutate(log_OTU1 = log10(OTU_00001 + 1)) %>%
  ggplot(aes(log_OTU1)) +
  geom_histogram(bins = 30)
```
:::

##### Task 3

Plot OTU 1 against OTU 2 as a scatter plot using `geom_point`.

Remember to add 1 before log10 transformation of both variables.

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

```{r, fig.width=4, fig.height=4}
#| out-width: "50%"
#| fig-align: "center"
counts %>%
  mutate(log_OTU1 = log10(OTU_00001 + 1)) %>%
  mutate(log_OTU2 = log10(OTU_00002 + 1)) %>%
  ggplot(aes(x = log_OTU1, y = log_OTU2)) +
  geom_point()
```
:::

##### Task 4

Look up the classifications of OTU 1 and OTU 2 in the taxonomy table.

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

```{r}
taxonomy %>% filter(OTU %in% c("OTU_00001", "OTU_00002"))
```

```{r}
taxonomy %>%
  filter(OTU %in% c("OTU_00001", "OTU_00002")) %>%
  select(OTU, Phylum, Family, Genus) 
```
:::

##### Task 5

Now make histograms and look up the taxonomy for the next thousand OTUs...

::: {.callout-note collapse="true" appearance="simple"}
##### SOLUTION:

Okay, that was a joke. 🤡

It is possible to make a thousand plots, because R is very good at repetitive tasks.

But, this would not be very useful, because you could not look at them all.

In the next section of this practical, we will explore smarter ways to analyse microbiota compositions.

:::
:::


## Next! ⏩

-   Click here: LINK TO NEXT PAGE

## Session info

<details>

```{r}
sessioninfo::session_info()
```

</details>
