---
title: "Practical 1A - R fundamentals"
subtitle: "2024 NUTRIM microbiome & metabolome workshop"
author: YOUR-NAME
date: last-modified
keep-md: false
embed-resources: true
format: html
theme: flatly
toc: true
---

## Instructions?

Follow the instructions online at:\
[david-barnett.github.io/2024-NUTRIM-microbiome/practical-1/web/practical1A-instructions.html](https://david-barnett.github.io/2024-NUTRIM-microbiome/practical-1/web/practical1A-instructions.html)

Some of the code and notes in this notebook are completed for you. The rest you will need to complete by following the the instructions.

-   Keep your notes and code in this document.

-   Check everything runs in order.

-   Take your time to understand each step.

-   Ask for help if confused.

### Using Quarto 📝

We will write code and notes in [Quarto](https://r4ds.hadley.nz/quarto){target="_blank"} `.qmd` documents, e.g. `practical-1A-notebook.qmd`

#### **Write some notes:**

You can write normal text for notes, e.g. your intentions, hypotheses, observations, etc.

To edit your Quarto notebooks in RStudio you can use "[Visual](https://quarto.org/docs/visual-editor/){target="_blank"}" editor mode, or the "Source" mode.

*Toggle between modes at the top left. Visual editor is easier for beginners, but Source mode shows how the Quarto file truly looks (it only contains [markdown](https://quarto.org/docs/authoring/markdown-basics.html "Markdown is plain text, with special characters to indicate headings, bold text, links etc"){target="_blank"} text, but RStudio adds formatting in Visual mode).*

#### **Write and run code:**

You can write code and comments in "chunks" - like the one below.

Run a chunk by clicking the green play button in the corner of the chunk

```{r}
# This is a comment, because it starts with the # symbol
print("This line is code, try to run it!") 
```

Insert a code chunk with `Ctrl`+`Alt`+`I` (Windows) or `Cmd`+`Option`+`I` (Mac).

```{r}
# You can also run code with keyboard shortcuts!
print("Run this one line with Ctrl + Enter, or Cmd + Enter on Mac")
print("or run a whole chunk with Ctrl/Cmd + SHIFT + Enter")
```

You can run all previous chunks by clicking the other button in the corner of the chunk.

## Load R packages 📦

Load all the required R packages.

```{r}

```

## Reading data 📂

Read metadata table from Excel file

```{r}
# give your objects short but informative names!
meta <- read_excel(path = here("data/papa2012/papa2012_metadata.xlsx"), na = "NA")
meds <- read_excel(here("data/papa2012/papa2012_metadata.xlsx"), sheet = "treatment") 
```

## Inspecting data

### Look at the tables 👀

Look at the metadata.

```{r}

```

### Summarise 📋

Check number of participants per diagnosis group.

```{r}
meta %>% count(diagnosis)
```

Check other category counts.

```{r}

```

Compute other summary statistics.

```{r}
meta %>% summarise(age_mean = mean(age_years), age_sd = sd(age_years))
```

Compute grouped summary statistics

```{r}

```

### Visualise 📊

Plot age histogram with base R

```{r}
#| fig-width: 6
#| fig-height: 3
meta$age_years %>% hist(breaks = 10, main = NULL, xlab = "Age")
```

Plot activity level bar chart with base R

```{r}
#| fig-width: 6
#| fig-height: 3
meta$activity %>% table() %>% barplot(xlab = "Disease Activity")
```

Plot age histogram with ggplot2

```{r}
#| fig-width: 6
#| fig-height: 3
ggplot(meta) +
  geom_histogram(aes(age_years), binwidth = 1) +
  facet_wrap(~case_control, ncol = 1) +
  theme_minimal()
```

Plot activity level bar chart with ggplot2

```{r}
#| fig-width: 6
#| fig-height: 3


```

## Joining data 🟧+️🟪

Join the main dataframe `meta` (containing most of the data about each patient) to the other dataframe `meds` (containing medications info for the IBD cases).

```{r}
all_meta <- left_join(meta, meds, by = 
```

Inspect the `all_meta` dataframe with `View(all_meta)`

## Mutating data 🧬

Data processing - modify variables and create new variables.

#### Base R 🤖

```{r}

```

#### dplyr::mutate 👽

```{r}

```

### Temporary mutation and plot

Temporarily mutate a dataframe and pipe the result into a plotting function.

```{r}
#| fig-width: 6
#| fig-height: 3
all_meta %>% 
  mutate(diagnosis = factor(diagnosis, levels = c("CD", "UC", "Other"))) %>% 
  ggplot(aes(diagnosis, age_years)) + geom_boxplot()
```

Check this did NOT persistently modify the diagnosis variable, it is still "character" class.

```{r}
class(all_meta$diagnosis) # no persistent changes, diagnosis is still character!
```

`mutate` dataframe and assign result to an object to persist the changes.

```{r}
#| fig-width: 6
#| fig-height: 3
all_meta <- all_meta %>% mutate(diagnosis = factor(diagnosis, c("CD", "UC", "Other")))
all_meta %>% ggplot(aes(diagnosis, age_years)) + geom_boxplot()
```

```{r}
class(all_meta$diagnosis) 
```

### More mutation

Clean up the medication history.

```{r}
table(all_meta$immunosuppression_level, useNA = "if")
```

```{r}
table(all_meta$medication, useNA = "if")
```

We have NAs for the medication for all the Controls. We know they had "none", so let's first indicate that. We will also replace Controls' immunosuppression_level NA values with "none".

```{r}

```

We now want logical variables indicating if the patient recently had antibiotics (abx), steroids, or other immunosuppressive drugs (imsp).

```{r}

```

For the character variables with more than two values, we convert them to factors, to encode our preferred ordering of their levels.

```{r}

```

## Save progress 💾

Save the `all_meta` dataframe in three formats: Excel, CSV, and RDS, all within a folder called "processed" inside the "data" folder.

First, create the "processed" directory.

```{r}

```

Write an excel file.

```{r}

```

ction from the `readr` package.

```{r}
write_csv(all_meta, here(output_path, "all_metadata.csv"))
```

RDS files are R's native format for saving single R objects. They preserve data types and structures, making them ideal for saving and loading a variety of R objects.

```{r}
saveRDS(all_meta, here(output_path, "all_metadata.rds"))
```

## Session info

```{r}
sessioninfo::session_info()
```
