---
title: "Practical 1B - microbiota data"
subtitle: "2024 NUTRIM microbiome & metabolome workshop"
author: YOUR-NAME
date: last-modified
keep-md: false
embed-resources: true
format: html
theme: flatly
toc: true
---

## Instructions

Follow the instructions online at: \
[david-barnett.github.io/2024-NUTRIM-microbiome/practical-1/web/practical1B-instructions.html](https://david-barnett.github.io/2024-NUTRIM-microbiome/practical-1/web/practical1B-instructions.html)

Some of the code and notes in this notebook are completed for you.

The rest you will need to complete by following the the instructions.

-   Keep your notes and code in this document.

-   Check everything runs in order.

-   Take your time to understand each step.

-   Ask for help if confused.

## Learning goals 🧠

-   Practice general R skills to inspect microbiota count data and taxonomy
-   Gain basic familiarity with specialist microbiome R packages: `phyloseq` and `microViz`
-   Learn to calculate, visualise, and interpret standard richness and diversity indices: Observed Richness and (Effective) Shannon diversity

## Load R packages 📦

```{r}
library(readxl)
library(here)
library(tidyverse)
```

## Read and inspect data 🔍

Read the metadata file.

```{r}
meta <- read_rds(file = here("data/papa2012/processed/all_metadata.rds"))
```

Read the count table.

```{r}
counts <- read_tsv(file = here("data/papa2012/papa2012_OTU_count_table.txt"))
```

```{r}
counts
```

#### **First challenge:**

Read the taxonomy table stored in `"data/papa2012/papa2012_taxonomy_table.txt"`

Call the object `taxonomy`

```{r}

```

### Taxonomy table

Now practice inspecting the taxonomy table by completing the following tasks:

#### Task 1

Check how many distinct genera there are.

```{r}

```

#### Task 2

How many OTUs are there in each Phylum?

```{r}

```

#### Task 3

What genera are in the phylum Actinobacteria?

```{r}

```

### OTU count table

Plot a histogram of OTU number 1.

```{r, fig.width=8, fig.height=2}
counts %>% ggplot(aes(OTU_00001)) + geom_histogram(binwidth = 5)
```

Attempt the following tasks to explore further!

#### Task 1

Use `filter()` to plot only the non-zero entries for OTU 1.

Try also using `+ scale_x_log10()` to transform the plot axis scale.

```{r, fig.width=8, fig.height=2}

```

#### Task 2

Use `mutate()` to create a temporary log-transformed variable `log_OTU1` and plot its distribution.

(Add 1 to all values first.)

```{r, fig.width=8, fig.height=2}

```

#### Task 3

Plot OTU 1 against OTU 2 as a scatter plot using `geom_point`.

Remember to add 1 before log10 transformation of both variables.

```{r, fig.width=4, fig.height=4}

```

##### Task 4

But what type of bacteria do these amplicon sequences belong to?

Look up the classifications of OTU 1 and OTU 2 in the taxonomy table.

```{r}

```

## Assemble a phyloseq 🛠️

Load specialist microbiota analysis packages

```{r}
library(phyloseq)
library(microViz)
```

Combining the three dataframes into one `phyloseq` object.

### OTU counts + taxonomy

phyloseq uses row or column names to match OTUs across the count and taxonomy tables.

*In addition, the taxonomy table must contain only taxonomic ranks in descending order, and the count table must be converted to a numeric matrix.*

Adjust the OTU count table.

```{r}
count_matrix <- counts %>% select(-sample) %>% as.matrix()
rownames(count_matrix) <- counts$sample
```

Adjust the taxonomy table.

```{r}

```

Combine the OTU and taxonomy tables into a phyloseq object: `ps`

```{r}

```

```{r}

```

### Adding sample metadata

phyloseq uses row names to match samples (across the OTU count table and the sample metadata).

```{r}
head(sample_names(ps))
```

In the phyloseq object the sample names have underscores, but they have hyphens in the sample metadata.

```{r}
meta
```

Fix that with the `stringr` package `str_replace` function.

```{r}

```

Now make them row names and check they all match.

```{r}

```

Add the metadata to the phyloseq object.

```{r}
sample_data(ps) <- meta_df
ps
```

## microViz data checks 🦠👁️

```{r}
# retrieve sample data as a tibble

```

```{r}
# get the OTU table, or part of it

```

```{r}
# get the taxonomy table

```

## Build some bar charts 📊

Each bar represents one sample, and we show the abundance data as proportions.

Bar chart with abundances aggregated by Phylum.

```{r, message=FALSE}
#| fig-width: 8
#| fig-height: 3


```

In this chart I see ....

Bar chart with abundances aggregated by Genus.

```{r, message=FALSE}
#| fig-width: 8
#| fig-height: 4


```

This plot is aggregated into genera, which provides more detail, but is harder to read, and we cannot give every genus a distinct colour.

We see that many samples contain a relatively large proportion of *Bacteroides* or *Faecalibacterium* but there is quite some further variation!

Update the `ps` phyloseq object with the default taxonomy table fixes.

```{r}
ps <- tax_fix(ps)
```

## Discover diversity 🌳🌲🌴

Calculating and visualising alpha diversity.

### Observed richness

Compute the observed richness of genera.

```{r}

```

### Shannon diversity

Calculate the Shannon diversity index at the level of Genus.

Also compute the exponentiated version, the effective number of genera, or effective Shannon index.

```{r}

```

### Diversity Distributions

Plot simple histograms of the richness and diversity values.

```{r}

```

### Interpreting diversity

## Save the phyloseq

```{r}

```

## Session info

```{r}
sessioninfo::session_info()
```
